---
title: "Tutorial to use TADreg R package"
author: "Raphael Mourad, Assistant Professor, Université Paul Sabatier"
date: "20/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is the main script to run the different functions:  
- SIM (Sparse Insulation Model) for identifying TAD borders / facilitators  
- DIM (Differential Insulation Model) for detecting differential TAD borders / facilitators  
- PIM (Prediction Insulation Model) for predicting Hi-C data with chromosomal rearrangements

## Load librairies
```{r load_lib}
suppressPackageStartupMessages(library(BSgenome.Mmusculus.UCSC.mm10))
suppressPackageStartupMessages(library(rtracklayer))
suppressPackageStartupMessages(library(GenomicRanges))
suppressPackageStartupMessages(library(HiTC))
suppressPackageStartupMessages(library(hicrep))

suppressPackageStartupMessages(library(Matrix))
suppressPackageStartupMessages(library(glmnet))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(mgcv))
suppressPackageStartupMessages(library(L0Learn))
suppressPackageStartupMessages(library(doMC))

suppressPackageStartupMessages(library(ggplot2))

suppressPackageStartupMessages(library(TADreg))
```

## For parallel computing
```{r parallel}
registerDoMC(cores=4)
```

## Create directories for results
```{r dir}
dir.create("results")
dir.create("results/SIM/")
dir.create("results/DIM/")
dir.create("results/PIM/")
```

### LOAD HIC DATA FOR SIM AND DIM RUNNING EXAMPLES 

## Parameters
```{r param}
res=25e3
reskb=paste0(res/1e3,"kb")
distMax=res*10
Chr="chr18"
```

## Cells
```{r cells}
cells=c("ES","CN")
```

## Loop over embryonic stem (ES) and cortical neuron (CN) cell lines
```{r load_data}
data(HiC_ES_mouse_chr18_50_60Mb_HTCexp) # load HTC_ES
data(HiC_CN_mouse_chr18_50_60Mb_HTCexp) # load HTC_CN
print(HTC_ES)
print(HTC_CN)
```

## Bins
```{r bins}
bin.GR=x_intervals(HTC_ES)
print(bin.GR)
```


### SPARSE INSULATION MODEL (SIM)  

Here we try to replicate results presented in the Figure 3 from the article. 
NB: results are slightly different from those presented in the Figure 3 from the article, because here we could not use the whole chromosome for computations since the file was too big to be deposited in an R package (5Mb limit). 

## Run Sparse Insulation Model (SIM) to identify TAD borders and facilitators from CN mouse cells. 
```{r SIM}
TADres=SIM(HTC_CN,distMax=distMax,penalty="L0")
print(TADres)
```

## Plot results from SIM
```{r plot_res_SIM}
x1=100
x2=300
posPlot=start(TADres)
coefPlot=TADres$beta
file_plotTAD=paste0("results/SIM/plot_TADborder_",Chr,"_",round(start(bin.GR[x1])/1e3),"kb_", end(bin.GR[x2])/1e3,"kb_",reskb,"_dmax",distMax/1e3,"kb.pdf")
pdf(file_plotTAD,12,4)
plot((posPlot/1e6)[x1:x2],coefPlot[x1:x2],type="l",xlab="Position (Mb)",ylab="SIM beta")
dev.off()
```

## Plot Hi-C heatmaps 
```{r plot_hic_map_SIM}
HTC_CNsub=extractRegion(HTC_CN,c(1,2),chr=Chr,from=start(bin.GR[x1]),to=end(bin.GR[x2]))
file_heatmapComp=paste("results/SIM/plot_HiC_CN_",Chr,"_",round(start(bin.GR[x1])/1e3),"kb_",end(bin.GR[x2])/1e3,"kb_",reskb,".pdf")
pdf(file_heatmapComp)
mapC(HTC_CNsub,log=T)
dev.off()
```


### DIFFERENTIAL INSULATION MODEL (DIM)

## Run Differential Insulation Model (DIM) to identify differential borders between CN and ES mouse cells 
```{r DIM}
TADdiffres=DIM(HTC_ES,HTC_CN,distMax)
coefDiff=TADdiffres$beta.diff
coefCond=cbind(TADdiffres$beta1,TADdiffres$beta2)
print(TADdiffres)
```

## Plot results at chr18:54888045-54990180 
```{r plot_res_DIM}
x1=100
x2=300
posPlot=(x1:x2)*res
coefPlot=coefCond[x1:x2,]
coefPlotDiff=coefDiff[x1:x2]
pvalPlotDiff=TADdiffres$padj.diff[x1:x2]
file_TADborderDiff=paste0("results/DIM/plot_TADborderDiff_",Chr,"_",round(start(bin.GR[x1])/1e3),"kb_",end(bin.GR[x2])/1e3,"kb_",reskb,"_d",distMax/1e3,"kb.pdf")
pdf(file_TADborderDiff,6,7)
par(mfrow=c(4,1))
par(mar=c(5,5,1,1))
minmax=c(min(coefPlot,coefPlotDiff),max(coefPlot,coefPlotDiff))
plot(posPlot/1e6,coefPlot[,1],ylim=minmax,type="l",xlab="Position (Mb)",
	ylab="Beta",col="red",main=cells[1])
plot(posPlot/1e6,coefPlot[,2],ylim=minmax,type="l",xlab="Position (Mb)",
	ylab="Beta",col="blue",main=cells[2])
plot(posPlot/1e6,coefPlotDiff,ylim=minmax,type="l",xlab="Position (Mb)",
	ylab="Beta",main="Differential borders")
plot(posPlot/1e6,-log10(pvalPlotDiff),type="l",xlab="Position (Mb)",
	ylab="-log10 adjusted p-value",main="Differential borders")
dev.off()
```

## Plot Hi-C heatmaps 
```{r plot_hic_map_DIM}
HTC_ESsub=extractRegion(HTC_ES,c(1,2),chr=Chr,from=start(bin.GR[x1]),to=end(bin.GR[x2]))
HTC_CNsub=extractRegion(HTC_CN,c(1,2),chr=Chr,from=start(bin.GR[x1]),to=end(bin.GR[x2]))
file_heatmapComp=paste("results/DIM/plot_HiC_",Chr,"_",round(start(bin.GR[x1])/1e3),"kb_",end(bin.GR[x2])/1e3,"kb_",reskb,".pdf")
pdf(file_heatmapComp)
mapC(HTC_ESsub,HTC_CNsub,log=T)
dev.off()
```


### PREDICTION INSULATION MODEL (PIM) 

## Parameters
```{r param_PIM}
res=10e3
reskb=paste0(res/1e3,"kb")
Chr="chr1"
```

## Loop over Hi-C files: WT (wild-type), DelB (Deletion B), DelBs (Deletion Bs) and InvF (Inversion F)
```{r load_data_PIM}
data(GSM2425366_CHiC_Epha4_WT_E115_MAPQ30_KRnorm_10kb_HTCexp) # HTC_WT
data(GSM2425367_CHiC_Epha4_DelB_E115_MAPQ30_KRnorm_10kb_HTCexp) # HTC_SVX
```

## Bins
```{r bin_PIM}
bin.GR=x_intervals(HTC_WT)
```

## Chromosomal rearrangement positions
```{r chr_rea}
SV.GR=GRanges("chr1",IRanges(c(76388978,76388978,74832836),c(78060839,77858974,75898707)))
typeSV=c("deletion","deletion","inversion")
SVX.GR=SV.GR[1]
typeSVX=typeSV[1]
```

## Parameters of PIM
```{r params_PIM}
distMax=res*ncol(intdata(HTC_WT))
model="glmlasso"
noise=1
output="asWT"
```

## Run Prediction Insulation Model (PIM) to predict Hi-C data after chromosomal rearrangement
```{r PIM}
HTC_PredSVX=PIM(HTC=HTC_WT,structVar.GR=SVX.GR,typeVar=typeSVX,
	output=output,model=model,distMax=distMax,parallel=T,noise=noise)
print(HTC_PredSVX)
```

## Correlations between predictions and observations
```{r cor_PIM}
count_SVX=as.vector(as.matrix(intdata(HTC_SVX)))
count_PredSVX=as.vector(as.matrix(intdata(HTC_PredSVX)))
dataCount=data.frame(SVX=count_SVX,PredSVX=count_PredSVX)
dataCount=dataCount[dataCount[,1]>0 & dataCount[,2]>0,]
corlogPredObs=round(cor(log(dataCount[,1]),log(dataCount[,2]),use="pairwise.complete.obs"),3)
corsPredObs=round(cor(count_SVX,count_PredSVX,method="spearman"),3)
SCCPredObs=round(compSCC(HTC_SVX,HTC_PredSVX),3)
```

## Plot heatmaps
```{r plot_heatmap_PIM}
Str=start(SVX.GR)
End=end(SVX.GR)
file_PredObs=paste0("results/PIM/heatmap_CHIC_PredObs_",Chr,"_",round(Str/1e3),"kb_",round(End/1e3),"kb.pdf")
pdf(file_PredObs)
mapC(HTC_SVX,HTC_PredSVX,log=T)
dev.off()
```

## Plot counts
```{r plot_counts_PIM}
file_plot_obspred_count=paste0("results/PIM/scatterplot_CHIC_PredObs_",Chr,"_",round(Str/1e3),"kb_",round(End/1e3),"kb.pdf")
pdf(file_plot_obspred_count,5,5)
d <- ggplot(dataCount, aes(SVX, PredSVX)) +
	geom_hex(bins=100) + scale_x_log10() + scale_y_log10() + coord_cartesian(xlim=c(1,500), ylim = c(1,500)) + 
	labs(title=paste0("Rlog=",corlogPredObs," Rs=",corsPredObs," SCC=",SCCPredObs))
print(d)
dev.off()
```



